Measurement of a dwarf galaxy rotational velocity from its Lyman-alpha line emission

### 2016-02-12 paper/letter.tex
first full draft of the main body of the paper

%\section{Introduction}

1. [General paragraph about the first galaxies and CDGs]

2. [General paragrph about the expected Lyman-alpha emission,
  LARS]. Almost fifty years ago \cite{PartridgePeebles} it was
realized that young galaxies could be detected through a stron
Lyman-$\alpha$ line emission.  This theoretical prediction was only
confirmed XX year later. 
Currently, Lyman Alpha Emitting galaxies are commonly measure in
surveys. 
The presence of this line provides confirmation of the distance of a
galaxy and also provides clues about the stellar population powering
the emission.

Lyman alpha emission is not exclusive of high redshift galaxies. It
does, require a low dust content to avoid complete extinction and an
source of ionizing radiation.  This has motivated the creation of
local Universe surveys to perform a detailed study of the Lyman-alpha
emission \cite{LARS}. A nearby simple permits the study of other
indicators that might be more difficult to obtain for distant galaxies
such as morphology, dust attenuation, neutral hydrogen contents and
ionization state.

3. [Observations require interpretation, common features and
  models]. The Lyman-$\alpha$ line is resonant. A yman-$\alpha$ photon
will follow a difussion-like process before escaping the galaxy or
being absorbed by dust. This makes it hard to interpret Lyman-alpha
line observations as the reulting profile is sensitive to the
dynamical, chemical and thermal conditions in the interstellar
medium. State-of-the-art models use mon.

4. [Rotation and the expected features.]
It has been shown that rotation also imprints an effect on the
Lyman-alpha morphology. 
The most important consequence of rotation is that spherical 
symmetry is broken.
The line morphology now depends on the viewing angle respect to the
rotation axis.  
For a line of sight perpendicular to the rotation axis the intensity
and the line center and the line width increase with rotational
velocity.  
When the rotational velocity is close to the half-line width of the
static line, the line becomes single peaked as it is observed in
\tol, a unique feature that other theoretical models find
impossible to reproduce.

5. [The charachteristics of the dwarf galaxy of interest as a puzzling
system]

\tol was first observed by ... it is a compact 
dwarf galaxy and does not have old stars.

The Ly$\alpha$ emission line was first observed in the \tol
galaxy by \cite{Thuan97}. It has two main important features which 
make this a very uncommun LAE. First it shown a symmetric profile 
which is , Second the Ly$\alpha$ line is not shifted with respect to 
the H$\beta$ line.  Blue compact Dwarf Galaxy

6. [The results of the fit + Physical picture using other results]

Figure 1. shows the observational data for \tol

with the overplot from
our best fit model from the full radiative transfer simulation. 
The parameters for the best fit are
$v_{max}=300$\kms, $\tau=1\times10^7$, $T=1.5\times 10^{4}$K and a
viewing angle $\theta<30$ degrees.

@@

%6. Comparison of these results against other observational 
Assuming spherical symmetry and a homogeneous gas distribution we
estimate the total neutral hydrogen mass to be on the order of
$M\approx m_H\tau^3\sigma^{-3}n^{-2}$, where $m_H$ is the mass a
Hydrogen's atom, $\tau$  the optical depth, $\sigma$ is the cross
section at the line's center and $n$ is the number density of neutral
Hydrogen atoms. 
For this system we estimate that for average values of $n=1\times
10^3$ the total hydrogen mass is $M\times 10^{14}$M$_{\odot}$. 
However, blind HI
surveys have put an upper limit in the neutral hydrogen mass of ???

7. [Possible interpretation of this results]

8. [Implications for outflow+rotation in existing samples.] Why are
not more examples of triple peaked emission? 
The ubiuquity of galaxy outflows overimposed to the rotation
transforms the line into the common double or single peaked
observations.

9. [Future outlook]

@@

From this limit and the rotational velocity of $300$ km/s and the
limit in the size $D$ we have limits of for the dynamical mass of
$2.1\times 10^{9}$M$_{\odot}$ $<M_D<$  $6.6\times 10^{9}$M$_{\odot}$.

which is at least $7$ to $25$ times larger than the H$I$ mass.

>>>>

The first generation of galaxies point to our cosmic corigins. Their
initial conditions and characteristiscs are the first page on the
evolution of galaxies such as the milky way that evolve to the present
time. In the standard big bang cosmology the only elemets that were
created in the nucleosynthesis where Hydrogen, Helium and
Lithium. Heavier elemets must hve been created in the internal
processes of stellar evolution. Therefore the first generation of
galaxies shoudl be metal free and rich in hydrogen. We haven detected
these galaxies yet. The best obsrevational proxy we have for them are
dwarf star forming galaxies with a low metallicity content. Compact
dwarf galaxies (CDGs) are the best observational example.

How can we develop a strategy to look for the first galaxies How can we recognize a first galaxy when we see one? 
Almost fifty years ago \cite{PartridgePeebles} realized that young
galaxies could be detected through a strong Lyman-$\alpha$ line
emission.  This theoretical prediction was only confirmed XX year
later.  Currently, Lyman Alpha Emitting galaxies are commonly measure
in surveys. 
The presence of this line provides confirmation of the distance of a
galaxy and also provides clues about the stellar population powering
the emission.

Lyman alpha emission, as suggested in the first paragraph, is not
exclusive of high redshift galaxies. Any galaxy with low metallicity
and dust content and ongoing star formation should present this line.  This has motivated the creation of
local Universe surveys to perform a detailed study of the Lyman-alpha
emission \cite{LARS}. A nearby simple permits the study of other
indicators that might be more difficult to obtain for distant galaxies
such as morphology, dust attenuation, neutral hydrogen contents and
ionization state.

The Lyman-$\alpha$ line is resonant. A yman-$\alpha$ photon
will follow a difussion-like process before escaping the galaxy or
being absorbed by dust. This makes it hard to interpret Lyman-alpha
line observations as the reulting profile is sensitive to the
dynamical, chemical and thermal conditions in the interstellar
medium. State-of-the-art models use mon.

It has been shown that rotation also imprints an effect on the
Lyman-alpha morphology. 
The most important consequence of rotation is that spherical 
symmetry is broken.
The line morphology now depends on the viewing angle respect to the
rotation axis.  
For a line of sight perpendicular to the rotation axis the intensity
and the line center and the line width increase with rotational
velocity.  
When the rotational velocity is close to the half-line width of the
static line, the line becomes single peaked as it is observed in
\tol, a unique feature that other theoretical models find
impossible to reproduce.

The Ly$\alpha$ emission line was first observed in the \tol
galaxy by \cite{Thuan97}. It has two main important features which 
make this a very uncommun LAE. First it has symmetric profile 
without any systemic displacement from the expected
line-center. Second it shows a triple peak, while the most common
observational feature corresponds either to a single or asymetric
double peak. 
This profile has escaped a clear theoretical understading based on
models that consider galaxies as spherical gas distributions with an
outflow. 
We have attempted an interpretation of the lya spectrum in \tol\ wihas
a central feature the bulk gas rotation without any outflow. This
model provides a complete explanation of \tol's puzzling feature.

Dots in Figure 1. represebt the observational data for \tol with the
overplot from our best fit model from the full radiative transfer
simulation.   The parameters for the best fit are
$v_{max}=300$\kms, $\tau=1\times10^7$, $T=1.5\times 10^{4}$K.
The optical depth and temperature can be translated into a column
density of $N_{HI}$. Our model is also able to constrain the angle
between the rotation axis and the observational line-of-sight to $\theta=65^{\circ}\pm$.
To constrain the model uncertainties we perform a Markov Chain Monte
Carlo calculation using an approximate analytical solution for the
lya spectrum in the bulk rotation case.

@@

Radio surveys have put an upper limit to the neutral hydrogen mass in
\tol of $M<2.65\times 10^{8}$ M$_{\odot}$. Together with our velocity
and column density results, under the assumption of spherical
symmetry, we can constrain the total dynamical mass of the HI region
to be in the range $2.1\times 10^{9}$M$_{\odot}$ $<M_D<$  $6.6\times
10^{9}$M$_{\odot}$ and the diameter of HI region should
have a diameter of $0.11 < D/\mathrm{kpc}<0.34$. This makes the
dynamical mass  at least $7$ to $25$ times larger than the H$I$ mass.

We consider three possibilities to contribute to this high
dynamical mass: stars, dark matter and a supermassive black
hole. The stellar mass can be constrained by measurements of the
stellar continuum, this can only represent $xx$M$_{\odot}$ of the total
contribution . Considering a standard variations for the dark matter
halo, it can only constributio to $xx$ M$_{\odot}$ in the region of
interest. A super massive black hole remains an open
possibility. Recent observations of ultra-compact dwarf galaxies
\cite{Seth2014} have confirmed the presence of super-massive black
holes containing $15\%$ of the total objects mass, suggesting that
there is a large population of undetected black holes.

A future observational test of this possibility in \tol\ would require
integral field unit measurements spatially resolving its spatial
extent. 
\tol\ spans a region of $\approx 4$ arcminutes. 
An instrument such as the Multi Unit Spectroscopic Explorer with its
nominal $0.2$ arcseconds spatial sampling over a $1.0$ arcminute field
in wide-field mode could provide a detailed mapping of different
ionization lines and infer a kinematic map.

Why are not more examples of triple peaked emission in observed
Lyman-alpha emitting gaalaxies?  
The ubiuquity of galaxy outflows overimposed to the rotation
transforms the line into the common double or single peaked
observations.  
Therefore, the absence of HI outflows in \tol\ is another special
feature that permitted a symmetric triple peaked lya line to emerge.

The implications of our finding for the study of lyman-alpha
emitters, including the first generation of galaxies are
manifold. First of all, it provides support to including rotation as a
feature to interpret observed spectra. Secondly, it provides a
motivation to explore under what conditions rotation dominated
dynamics are expected in comparison with outflow dominated
dynamics. Finally, if the hypothesis of a supermassive black hole in
\tol\ proves to be consistend with future observational kinematic
maps, it would provide a new information to test and probe theories on
the coevolution of galaxies and black holes.

@@

From this limit and the rotational velocity of $300$ km/s and the
limit in the size $D$ we have limits of for the dynamical mass of
$2.1\times 10^{9}$M$_{\odot}$ $<M_D<$  $6.6\times 10^{9}$M$_{\odot}$.
which is at least $7$ to $25$ times larger than the H$I$ mass.
